---
title: "SAM vs Stunting Analysis - Part 1"
author: Monica N Mweetwa
date: 30/12/2024
---
Set working directory
```{r setup}
   #knitr::opts_knit$set(root.dir = normalizePath("~/Dropbox/BEECH_16s_Analysis/Stunting_SAM")) 
   knitr::opts_knit$set(root.dir = normalizePath("C:/Users/Monica/Dropbox/BEECH_16s_Analysis/Stunting_SAM")) 
```

load data
```{r}
library(phyloseq)
library(ggpubr)
library("ggplot2")
library(tableone)
library(tidyverse)
library(microViz)
library(pals) #for color palette
load("Data/RData/phyloseq_dataset_species.RData") #ps0
```

#1. Visualization of relative abundance of top 15 genera in both datasets
```{r}
otu_palette <- c(unname(alphabet2(n=15)), "lightgray")

#Genus level
a <- ps0 %>%
  ps_mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>%
  tax_fix() %>% 
  comp_barplot(tax_level = "Genus",
               label = "PID",
               n_taxa = 15) +
  coord_flip() +
  facet_grid(
    rows = vars(mal.type),
    scales = "free_y", # this only frees y scale per row in grid faceting
    space = "free_y" # allows bars to be same size by freeing facet heights
  ) +
  theme(axis.text.y = (element_blank()),
        axis.ticks.y = (element_blank()))

ggsave("Output/RelAbund/RelAbund_gen.png", plot = a, dpi = 600, device = "png", width = 8, height = 8, units = "in")

#get relative abundance numbers
ps0_gen <- ps0 %>%
  ps_mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>% 
  tax_fix() %>%
  tax_transform("compositional", rank = "Genus") %>%
  ps_melt() %>%
  group_by(mal.type, OTU) %>%
  summarize(N = n(),
            mean_val = round((mean(Abundance, na.rm = T))*100,5),
            SD_val = round((sd(Abundance, na.rm = T))*100,5)) 
write_csv(ps0_gen, "Output/RelAbund/Gen_abund.csv")

#Phylum level
a2 <- ps0 %>%
  ps_mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>%
  tax_fix() %>% 
  comp_barplot(tax_level = "Phylum",
               label = "PID",
               n_taxa = 15) +
  coord_flip() +
  facet_grid(
    rows = vars(mal.type),
    scales = "free_y", # this only frees y scale per row in grid faceting
    space = "free_y" # allows bars to be same size by freeing facet heights
  ) +
  theme(axis.text.y = (element_blank()),
        axis.ticks.y = (element_blank()))

ggsave("Output/RelAbund/RelAbund_phy.png", plot = a2, dpi = 600, device = "png", width = 8, height = 8, units = "in")

#get relative abundance numbers
ps0_phy <- ps0 %>% 
  ps_mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>%
  tax_fix() %>%
  tax_transform("compositional", rank = "Phylum") %>%
  ps_melt() %>%
  group_by(mal.type, OTU) %>%
  summarize(N = n(),
            mean_val = round((mean(Abundance, na.rm = T))*100,5),
            SD_val = round((sd(Abundance, na.rm = T))*100,5)) 
write_csv(ps0_phy, "Output/RelAbund/Phy_abund.csv")
```

#2. Alpha diversity

```{r}
#Calculate Shannon
ad = estimate_richness(ps0, measures = "Shannon")
ad <- rownames_to_column(ad, "SampleID")

#Calculate Faith's PD estimate
library(btools)
pd <- estimate_pd(ps0)
pd <- rownames_to_column(pd, "SampleID")

#Combine measures
AlphaDiv = merge(pd, ad, by = "SampleID")

add <- data.frame(sample_data(ps0))
add <- rownames_to_column(add, "SampleID2")

add2 = merge(add, AlphaDiv, by = "SampleID")
```

Relationship between alpha diversity (Faith's PD) - Clinical features.

a. Histograms - Check distribution of variables
```{r}
hist(add2$PD)
hist(add2$Shannon)
hist(add2$AgeMonths_endo)
hist(add2$WAZ_endo)
hist(add2$WHZ_endo)
hist(add2$LAZ_endo)
hist(add2$GastricPh_endo)
hist(add2$GLP.2_endo)
hist(add2$I.FABP_new)
hist(add2$LPS_endo)
hist(add2$sCD14_new)
hist(add2$VH)
hist(add2$CD)
hist(add2$ESA)
```


Spearman correlation - Clinical features 
```{r}
add3 <- add2 %>%
  rename("Age (Months)" = "AgeMonths_endo",
         "WAZ" = "WAZ_endo",
         "LAZ" = "LAZ_endo",
         "WHZ" = "WHZ_endo" ,
         "GLP2" = "GLP.2_endo",
         "iFABP" = "I.FABP_new",
         "LPS" = "LPS_endo",
         "sCD14" = "sCD14_new",
         "Gastric pH" = "GastricPh_endo",
         "HIV status" = "hiv_char")

#Faith's PD
spearman_plots <- add3 %>% 
  mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>%
  pivot_longer(names_to = "clin_vars", values_to = "vals", cols = c(`Age (Months)`, WAZ,
                                                                    WHZ, LAZ,GLP2,
                                                                    iFABP,LPS, sCD14, VH,
                                                                    CD, ESA, `Gastric pH`)) %>%
  filter(!is.na(vals)) %>%
  filter(vals != 0) %>%
  ggplot(aes(x = vals, y = PD)) +
  geom_point() +
  labs(x = "Clinical Features",
       y = "Faith's PD") +
  stat_cor(method = "spearman", ) +
  geom_smooth(method = "lm") +
  facet_grid(mal.type ~ clin_vars,scales="free",margin="mal.type",drop = TRUE,switch="both")
ggsave("Output/AlphaDiv/AlphaDiv_spearman_pd.png", plot = spearman_plots, dpi = 600, device = "png", width = 22, height = 6, units = "in")

#Shannon
spearman_plots2 <- add3 %>% 
  mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>%
  pivot_longer(names_to = "clin_vars", values_to = "vals", cols = c(`Age (Months)`, WAZ,
                                                                    WHZ, LAZ,GLP2,
                                                                    iFABP,LPS, sCD14, VH,
                                                                    CD, ESA, `Gastric pH`)) %>%
  filter(!is.na(vals)) %>%
  filter(vals != 0) %>%
  ggplot(aes(x = vals, y = Shannon)) +
  geom_point() +
  labs(x = "Clinical Features",
       y = "Shannon's index") +
  stat_cor(method = "spearman", label.y = 4) +
  geom_smooth(method = "lm") +
  facet_grid(mal.type ~ clin_vars,scales="free",margin="mal.type",drop = TRUE,switch="both")
ggsave("Output/AlphaDiv/AlphaDiv_spearman_shannon.png", plot = spearman_plots2, dpi = 600, device = "png", width = 22, height = 6, units = "in")
```

Boxplots for categorical variables - Wilcoxon's test 
```{r}
# malnutrition type 
colors = c("SAM" = "#336633", "Stunting" = "#CC6600")
my_comparisons = list(c("SAM", "Stunting"))
b <- add2 %>%
  mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>%
  ggplot(aes(x = mal.type, y = Shannon, fill = mal.type)) + 
  geom_boxplot(outlier.shape = NA, size = 0.8, width = 0.8, na.rm = TRUE) +
  theme_bw() + 
  scale_fill_manual(values = colors) +
  stat_compare_means() + #statistical comparison
  labs(x = " ", 
       y = "Shannon Index")  +
  theme(axis.text.x = element_text(vjust = 0.5),
        plot.title = element_text(size = 25, hjust = 0.5),
        legend.position = "none")
ggsave("Output/AlphaDiv/AlphaDiv_wilcoxon_shannon.png", plot = b, dpi = 600, device = "png", width = 4, height = 4, units = "in")
b2 <- add2 %>%
  mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>%
  ggplot(aes(x = mal.type, y = PD, fill = mal.type)) + 
  geom_boxplot(outlier.shape = NA, size = 0.8, width = 0.8, na.rm = TRUE) +
  theme_bw() + 
  scale_fill_manual(values = colors) +
  stat_compare_means() + #statistical comparison
  labs(x = " ", 
       y = "Faith's PD")  +
  theme(axis.text.x = element_text(vjust = 0.5),
        plot.title = element_text(size = 25, hjust = 0.5),
        legend.position = "none")
ggsave("Output/AlphaDiv/AlphaDiv_wilcoxon_pd.png", plot = b2, dpi = 600, device = "png", width = 4, height = 4, units = "in")
```

3. Beta diversity
```{r}
source("Code/Functions.R")
#Combine at genus level
ps0.gen <- tax_glom(ps0, taxrank = "Genus", NArm = TRUE)
#Rename taxa with genus names
genus.names <- make.names(tax_table(ps0.gen)[,'Genus'], unique = TRUE)
taxa_names(ps0.gen) <- genus.names

#Transform to relative composition
ps0.gen_comp <- ps0 %>%
  ps_mutate(mal.type = ifelse(study == "EE children", "SAM", "Stunting")) %>%
  tax_fix() %>%
  tax_transform("compositional", rank = "Genus")
  
#Permanova analysis @ genus level
lm_mal <- permanova_func(equation = dist ~ mal.type, data = ps0.gen_comp)
write.csv(lm_mal, "Output/BetaDiv/BetaDiv_mal.csv")
rm(list = ls()[grep("^lm", ls())])

#Vizualize the first 2 principal components
library(microViz)
c <- ps0.gen %>%
  tax_fix() %>%
  tax_transform("compositional", rank = "Genus") %>%
  dist_calc("bray") %>%
  ord_calc(method = "PCoA") %>%
  ord_plot(color = "mal.type", plot_taxa = 1:10, size = 5, expand = TRUE,
           tax_lab_style = tax_lab_style(type = "text", size = 4, check_overlap = TRUE)) +
  scale_colour_manual(values = colors) +
  labs(color = "Malnutrition Class") +
  theme_bw() +
  #annotate("text", x = 2, y = -2.5, size = 3,
           #label = paste("PERMANOVA Test: R2 = 0.075, p = 0.001")) +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(subtitle = "PERMANOVA Test: R2 = 0.065, p = 0.001")
ggsave("Output/BetaDiv/BetaDiv.png", plot = c, dpi = 600, device = "png", width = 8, height = 8, units = "in")
```

Combine plots
```{r}
library(ggpubr)

ggarrange(a, ggarrange(ggarrange(b,b2,nrow = 1),
                       c, nrow = 2, labels = c("B.", "C.")),
          nrow = 1, labels = c("A."))
ggsave("Output/combined_plot.png", plot = last_plot(), dpi = 600, device = "png", 
       width = 12, height = 8, units = "in")
```

#DONE